Supporting Information for\\ ``A heterogeneous landscape of selection and epistasis in protein-coding genes''

### 2022-03-25 manuscript/supplement.tex
getting closer..

\usepackage[margin=1in]{geometry}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{xspace}
\usepackage{amsmath,amssymb}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths

@@

\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{equation}{0}

\small
\tableofcontents
\normalsize
\newpage

@@

\subsection{Drift, mutation, recombination, and moment closure}\label{drift-mutation-recombination-and-moment-closure}

\subsubsection{Drift}\label{drift}

@@

\subsubsection{Recombination}\label{recombination}

@@

\subsubsection{Mutation}\label{mutation}

@@

\subsection{Selection}\label{selection}

@@

\subsection{Validation against simulation}\label{validation-against-simulation}

\subsection{Background selection and associative overdominance}\label{background-selection-and-associative-overdominance}

\subsection{Frequency dependence of signed LD}\label{frequency-dependence-of-signed-ld}

\section{Data analysis}\label{data-analysis}

\subsection{DFE for missense and LOF variants}\label{dfe-for-missense-and-lof-variants}

@@

\subsection{Multinucleotide mutations and positive LD between linked synonymous variants}\label{multinucleotide-mutations-and-positive-ld-between-linked-synonymous-variants}

@@

MNM events can be easily incorporated into the moment system with a simple
adjustment to the mutation operator. Instead of all mutations occurring
independently in haplotypes with mutations already segregating at the other
locus, some fraction of new mutations could instead occur spontaneously and
create a new pair of mutations with initial counts \(n_{AB} = 1\) and \(n_{ab} = n-1\).

Here, I fit a simple exponential model for the fraction of new mutations at a
given distance that arose through a MNM event, so that \(P(MNM | d) = Ae^{-\lambda d}\), where \(d\) is the distance separating pairs of mutations. I
considered all synonymous mutations within genes in the MSL population and used
the same population size history model as inferred in the DFE section above for
a demographic control. This left two parameters to be fit, \(A\) and \(\lambda\),
which I fit to the binned decay curve of \(\sigma_d^1\). I needed to assume an
average per-base recombination rate \(r\) across gene regions, and tested a
number of values between \(10^{-9}\) and \(2\times 10^{-8}\). The optimization was
insensitive to the chosen value of \(r\), because the decay of positive LD occurs
rapidly. For any plausible value of \(r\), \(\sigma_d^1\) decays to zero well
before distances between pairs have scaled recombination rates \(\rho=4 N_e r d \sim 1\), and expected statistics for \(\rho \ll 1\) vary only negligibly.

In fitting the LD decay of \(\sigma_d^1\), the best fit parameters were
\(A=0.132\), and \(\lambda=0.0103\). An exponential scaling of \(0.01\) implies that
the vast majority of new mutations \emph{pairs} do not occur via MNMs for distances
greater than 200 bp, though a substantial fraction (\(10-15\%\)) occur via MNMs
for very tightly linked loci with distances on the order \(0-50\) base pairs. It
is important to note that this does not mean that \(10-15\%\) of new mutations
occur via MNMs, since this fraction is conditioned on two mutations occurring
at short distances.

\subsection{Grouping Thousand Genomes populations based on clustering}
\label{grouping-thousand-genomes-populations-based-on-clustering}

@@

\addcontentsline{toc}{section}{References}
\bibliographystyle{genetics}
\bibliography{paper}

@@

\addcontentsline{toc}{section}{Supporting tables}
\section*{Supporting tables}

@@

\addcontentsline{toc}{section}{Supporting Figures}
\section*{Supporting figures}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/jackknife}
    \caption{
        \textbf{Accuracy of the jackknife approximation and runtime} Small
        sample sizes can lead to large error in the moment-closure
        approximation for larger recombination distances or selection
        coefficients. Generally, the jackknife approximation breaks down for
        recombination rates greater than \(\rho\approx30\). While increasing
        sample size leads to more accurate solutions, it comes at the cost of
        both increased runtime and memory usage. Most analyses in this paper
        used sample sizes between 40 and 80.
    }
    \label{fig:jackknife}
\end{figure}

@@

@@

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/msl_mnms}
    \caption{
        \textbf{Optimization of fraction of new mutations arising via
        multinucleotide mutations by distance.} A simple exponential function was fit
        to describe the probability that a pair of mutations arose through a MSM event
        at a given distance \(d\), as \(Ae^{-\lambda d}\). Across all recombination rates
        tested, the best fit parameters were \(A=0.13\) and \(\lambda=0.010\).
    }
    \label{fig:mslmnms}
\end{figure}

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{xspace}
\usepackage{amsmath,amssymb}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
\usepackage[numbib]{tocbibind} % makes the references have a section number
\renewcommand{\bibname}{Supporting references}

@@

\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thesection}{S\arabic{section}}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{equation}{0}
\setcounter{section}{0}

\tableofcontents
\newpage

@@

\subsection{Drift, mutation, recombination, and moment closure}

\subsubsection{Drift}

@@

\subsubsection{Recombination}

@@

\subsubsection{Mutation}

@@

\subsection{Selection}

@@

\subsection{Validation against simulation}

To assess the accuracy of the numerical approximation, I carried out discrete
two-locus simulations over a range of recombination rates and selection
parameterizations. Such discrete simulations track allele and haplotype
frequencies in a population of a given size, with offspring generations
constructed via multinomial sampling from the haplotype frequencies in the
parental generation. Because we only track the two focal loci, these
simulations do not contain any additional effects of linked selection (see
Section~\ref{sec:bgs} for simulations with many linked loci).

Simulations had diploid population sizes of 5000 and I drew haploid sample
sizes of 50 (25 diploids), with recombination rates varying from \(0\) to
\(0.001\) (or \(\rho=0-10\)). I tested selection coefficients of \(s\) of 0,
\(-2\times10^{-5}\), \(-2\times10^{-4}\), and \(-2\times10^{-3}\) (or
\(\gamma=0\), \(-0.2\), \(-2\), and \(-10\)). Epistasis was set to
\(\epsilon=0\), \(-0.5\), and \(0.5\). Conditional distributions of the
two-locus sampling distribution show good agreement between numerical solutions
and simulations, and residuals do not show any directional bias
(Figures~\ref{fig:validation}--\ref{fig:last_validation}).

To assess the accuracy under dominance, I simulated under the
\citet{Roze2021-cf} scenarios (\(N=1000\), \(r=0.0001\) and \(0.001\), and
\(sh=-0.0001\), \(-0.0004\), \(-0.001\), and \(-0.01\)). Note that \(sh=-0.01\)
is extremely strong selection, especially for small \(h\), and \texttt{moments}
was unstable in that parameter regime. For weak and moderate selection, there
good agreement between simulated data and \texttt{moments} solutions
(Figure~\ref{fig:dominance}D, E).

\subsection{Background selection and associative overdominance}\label{sec:bgs}

To explore the effects of many linked selected loci on patterns of LD, I used
\texttt{fwdpy11} \citep{Thornton2014-pn,Thornton2019-qc} to simulate large
genomic segments with many selected mutations. Each simulation had population
size of \(N=1000\), a sequence length of 1Mb, with mutation and recombination
rates both \(u=r=10^{-8}\). I considered four scenarios for selected mutaions:
additive selection \(h=0.5\) with \(s=-0.0005\) and \(s=-0.01\) (\(\gamma=-1\)
and \(-20\)), and recessivity \(h=0.0\) or \(h=0.1\) with the same selection
coefficients. These sets of parameters should give rise to strong effects of
linked selection, as there are many cosegregating, tightly linked selected
variants. For each setting, I examined site-frequency spectra and LD decay for
both selected and neutral mutations across the region.

For each simulation, I ran a large number of replicates (100 independent
replicates, and within each replicate, I sampled 25 diploid individuals every
100 generations for a total of 20,000 samples from each replicate, after
burn-in of \(100N\) generations). Additive selection caused minor distortions
to the SFS, and reduced observed \(\sigma_d^1\) and \(\sigma_d^2\) in the case
of weak selection (\(\gamma=-1\)) (Figure~\ref{fig:bgs1}). LD was largely
unaffected in the stronger selection scenario, consistent with intuition that
interference (including interference from additional linked sites) is most
noticeable for \(s\approx1/N\) (Figure~\ref{fig:bgs2}).

Larger deviations from two-locus expectations were observed for recessive
mutations. For recessive weakly deleterious mutations (Figure~\ref{fig:bgs3}),
there is a large excess of neutral and selected mutations at common frequencies
(due to associative overdominance \citep{Zhao2016-bb}). \(\sigma_d^2\) for
neutral and selected mutations is also increased over two-locus expectations.
Stronger recessive mutations (Figure~\ref{fig:bgs4}) also show distorted allele
frequencies, though the distortion to LD is not as strong as observed for weak
recessive variants.

\subsection{Frequency dependence of signed LD}

\begin{itemize}
    \item Why did we do it
    \item How did we do it.. quite simple
\end{itemize}

\section{Data analysis}

\subsection{DFE for missense and LOF variants}

@@

\subsection{Multinucleotide mutations and positive LD between linked synonymous variants}

@@

MNM events can be easily incorporated into the moment system with a simple
adjustment to the mutation operator. Instead of all mutations occurring
independently in haplotypes with mutations already segregating at the other
locus, some fraction of new mutations could instead occur spontaneously and
create a new pair of mutations with initial counts \(n_{AB}=1\) and
\(n_{ab}=n-1\). We can partition $\Psi_n$ into pairs of mutations due to MNM
events (\(\Psi_n^{MNM}\)) and those due to sequential mutation events
(\(\Psi_n^{ISM}\)), so that \(\Psi_n=\Psi_n^{MNM}+\Psi_n^{ISM}\). For given
scaled mutation rates \(\theta=4N_e\mu\), \(\Psi_n^{ISM}\propto\theta^2\) and
\(\Psi_n^{MNM}\propto p_{MNM}\theta\), where \(p_{MNM}\) is the probability
that a mutational event causes a multinucleotide mutation. This difference in
scaling means that \(\Psi_n\) is sensitive to both the overall mutation rate
\(\theta\) as well as the probability that a mutation is a MNM at a given
distance.

\subsubsection{Optimization details}

Here, I fit a simple exponential model for the fraction of new mutations at a
given distance that give rise to a MNM event, so that \(p_{MNM,d} = P(MNM | d)
= Ae^{-\lambda d}\), where \(d\) is the distance separating pairs of mutations
in base pairs. I considered all synonymous mutations within genes in the MSL
data and used the same population size history model as inferred in the DFE
section above for a demographic control. This left two parameters to be fit,
\(A\) and \(\lambda\), which I fit to the binned decay curve of \(\sigma_d^1\)
using the midpoint of each bin as the distance for that bin. I needed to assume
an average per-base recombination rate \(r\) across gene regions, and tested a
number of values between \(10^{-9}\) and \(2\times 10^{-8}\). Optimization was
insensitive to the chosen value of \(r\), because the decay of positive LD
occurs rapidly. For any plausible value of \(r\), \(\sigma_d^1\) decays to zero
well before distances between pairs have scaled recombination rates \(\rho=4
N_e r d \sim 1\), and expected statistics for \(\rho \ll 1\) vary negligibly.

To perform optimization, I assumed a Guassian likelihood function within each
bin using variances estimated via bootstrap replicates from sampling genes with
replacement. A composite likelihood was computed by taking the product of
likelihoods across bins. I used the estimate of \(Ne\approx12,300\) from the
DFE optimization using MSL data, a coding mutation rate of
\(\mu=1.33\times10^{-8}\) \citep{Karczewski2020-le}, and the fraction of coding
mutations that result in synonymous mutations was assumed to be (assuming a
ration of nonsynonymous mutations to synonymous mutations of 2.5:1). Thus, the
scaled mutation rate was \(\theta=4\times
N_e\times\mu\times\frac{1}{3.5}\approx1.87\times10^{-4}\).

\subsubsection{MNM optimization results and discussion}

In fitting the LD decay of \(\sigma_d^1\), the best fit parameters were
\(A=2.44\times10^{-5}\), and \(\lambda=0.00919\). Thus, only a small fraction
of new mutations were inferred to cause MNM events (\(\ll1\%\)), and
\(1/\lambda\approx 100\) means that the proportion of mutation events that
cause MNM decays quickly within a few hundred base pairs.

However, because of the differences in scaling of \(\Psi_n^{ISM}\) and
\(\Psi_n^{MNM}\) (with the square of \(\theta\) vs linearly as
\(p_{MNM}\theta\)), the number of \emph{observed} pairs of mutations at a given
distance that arose due to a MNM event is much larger. A rough estimate given
the assumed mutation rate and inferred parameters is that \(p_{MNM}\theta /
\theta^2 \approx 10\%\) of pairs of synonymous mutations observed at
\(d\approx0\) are due to MNM events, roughly \(p_{MNM}e^{-\lambda 100}\theta /
\theta^2 \approx 4\%\) at \(d=100\) base pairs, and only about
\(p_{MNM}e^{-\lambda 400}\theta / \theta^2 \approx 0.25\%\) at \(d=400\) base
pairs. This inference is very sensitive to the assumed mutation rate and the
proportion of mutations that cause synonymous vs nonsynonymous variants, and
future work will be needed to refine these estimates.

\subsection{Grouping Thousand Genomes populations based on clustering}

@@

\bibliographystyle{genetics}
\bibliography{paper}

@@

\section{Supporting tables}

@@

\section{Supporting figures}

\textbf{all supporting figures need captions}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/jackknife}
    \caption{
        \textbf{Accuracy of the jackknife approximation and runtime.}
        Small sample sizes can lead to large error in the moment-closure
        approximation for larger recombination distances or selection
        coefficients. Generally, the jackknife approximation breaks down for
        recombination rates greater than \(\rho\approx30\). While increasing
        sample size leads to more accurate solutions, it comes at the cost of
        both increased runtime and memory usage. Most analyses in this paper
        used sample sizes between 40 and 80.
    }
    \label{fig:jackknife}
\end{figure}

\begin{figure}[ht!]
    \centering
    %\includegraphics{}
    \caption{\textbf{Validation figures.}}
    \label{fig:validation}
\end{figure}

\begin{figure}[ht!]
    \centering
    %\includegraphics{}
    \caption{\textbf{Validation figures.}}
    \label{fig:last_validation}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\textwidth]{../figures/data_within_between}
    \caption{
        \textbf{Within vs between.}
    }
    \label{fig:between_domains}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/msl_mnms.v2}
    \caption{
        \textbf{Optimization of fraction of new mutations arising via
        multinucleotide mutations by distance.} A simple exponential function was fit
        to describe the probability that a pair of mutations arose through a MSM event
        at a given distance \(d\), as \(Ae^{-\lambda d}\). Across all recombination rates
        tested, the best fit parameters were \(A=0.11\) and \(\lambda=0.0085\).
    }
    \label{fig:mslmnms}
\end{figure}
